---
title: "R_Assignment"
author: "Kavi R. Acharya"
date: "March 19, 2021"
output:
  word_document: default
  pdf_document: default
  html_document: default

## Part I:  Unix assignment in R

Loading the tidyverse package:
```{r}
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)

```

## Downloading the Fang et al and snp files

```{r} 
# Downloading files directly from the github repository 
snp<-read_tsv("https://raw.githubusercontent.com/EEOB-BioData/BCB546-Spring2021/main/assignments/UNIX_Assignment/snp_position.txt")

Fang<-read_tsv("https://raw.githubusercontent.com/EEOB-BioData/BCB546-Spring2021/main/assignments/UNIX_Assignment/fang_et_al_genotypes.txt")

```
## Data Inspection

we can inspect the data fang et al as well as snp datafile using following 

```{r}
# Load data:
str(snp)
unique(genotypes$Group)
unique(snp$Chromosome)
nrow(snp)
ncol(snp)
```



### Data Processing

At first we need to rearrange the snp file. SNP file is rearragned so that Chromosome is kept at second column  followed by the position. Then the file is transposed.

```{r}
# Rearranging the snp file
snp<-snp[c(1,3,4,2,5:15)]

#Creating subsets of Maize and Teosinte genotype files

Maize.genotype<-Fang %>% filter(Group=="ZMMIL"|Group=="ZMMLR"|Group=="ZMMMR") 

Teosinte.genotype<-Fang %>%filter(Group=="ZMPJA"|Group=="ZMPIL"|Group=="ZMPBA")


```

### Now, we transpose the subsets and merge. 
```{r}

Maize.genotype<-as.data.frame(t(Maize.genotype), stringsAsFactors = F)
Teosinte.genotype<-as.data.frame(t(Teosinte.genotype), stringsAsFactors = F)

#Here, we are transposing column name intp row name. Then, a new row name is created under SNP_ID. The first row is converted into row name.At last, first three columns are removed.
SNP_ID <- rownames(Maize.genotype)
rownames(Maize.genotype) <- NULL
Maize.genotype <- cbind(SNP_ID,Maize.genotype,stringsAsFactors = FALSE)
names(Maize.genotype)<-  c("SNP_ID",Maize.genotype[1,-1])
Maize.genotype <- Maize.genotype[-c(1,2,3), ]


SNP_ID <- rownames(Teosinte.genotype)
rownames(Teosinte.genotype) <- NULL
Teosinte.genotype <- cbind(SNP_ID,Teosinte.genotype,stringsAsFactors = FALSE)
names(Teosinte.genotype)<-  c("SNP_ID",Teosinte.genotype[1,-1])
Teosinte.genotype <- Teosinte.genotype[-c(1,2,3), ]


```

Now, we are going to merge the transposed genotype file with snp files
```{r}
Merged.maize<-merge(snp, Maize.genotype, by="SNP_ID")
Merged.teosinte<-merge(snp, Teosinte.genotype, by="SNP_ID")
```
We now subset the merged dataset based on Chromosome number, sort it based on position and export it in the form of csv files

```{r}
#Here, the as.numeric function will give us an error message when there is "unknow/multiple" value. We will replace such value by NA. New column Pos is created and the data is sorted numerically. After sorting, Pos is removed. 

for(i in c(1:10)){
  data<-Merged.maize %>% filter(Chromosome==i)%>% mutate(Pos=as.numeric(Position))%>%arrange(Pos)
  data$Pos<-NULL
  write.csv( data,paste0("Maize_Chromo_",i,"_ascending.csv"), row.names = F)
  data<-Merged.maize%>% filter(Chromosome==i)%>% mutate(Pos=as.numeric(Position))%>%
    arrange(-Pos)
  data$Pos<-NULL
  data[data=="?/?"]<-"-/-"
  write.csv( data,paste0("Maize_Chromo_",i,"_descending.csv"), row.names = F)
  
  data<-Merged.teosinte %>% filter(Chromosome==i)%>% mutate(Pos=as.numeric(Position))%>%arrange(Pos)
  data$Pos<-NULL
  write.csv( data,paste0("Teosinte_Chromo_",i,"_ascending.csv"), row.names = F)
  data<-Merged.teosinte%>% filter(Chromosome==i)%>% mutate(Pos=as.numeric(Position))%>%
    arrange(-Pos)
  data$Pos<-NULL
  data[data=="?/?"]<-"-/-"
  write.csv( data,paste0("Teosinte_Chromo_",i,"_descending.csv"), row.names = F)
  
}



```




# Part II

Data Visualization.

### SNPs per chromosome
We are going to visualize the  number of possible polymorphism in each chromosome.


```{r}
library(ggplot2)

ggplot(data = snp[!is.na(as.numeric(snp$Chromosome)),]) +   geom_bar(mapping = aes(x = as.numeric(Chromosome), fill=Chromosome)) + scale_x_discrete(limit=c(1:10))+ labs(x = "Chromosome number", y="No. of polymorphism position") 
```

```{r}
#Traforming the data using pivot_longer. 

Pivot<-Fang %>% pivot_longer(!c(Sample_ID, JG_OTU, Group),names_to="SNP_ID",values_to= "Base")
Pivot<-merge(Pivot, snp, by="SNP_ID")

ggplot(data = Pivot[!is.na(as.numeric(Pivot$Chromosome)),]) +   geom_bar(mapping = aes(  as.numeric(Chromosome), fill=Group)) + scale_x_discrete(limit=c(1:10))+ labs(x = "Chromosome number", y="No. of SNPS")


```

```{r}
#To see polymorphism in each group

Group_graph <- Pivot %>% 
  mutate(Chromosome=as.numeric(Chromosome)) %>% #This will mutate Chromosone number into numeric value
  select(Group, SNP_ID, Chromosome, Base) %>%  #this function will Select required fields
  filter(Chromosome %in% c(1:10)) %>%  #Filters chromosome 1 to 10
  filter(Base!="?/?")%>%   #Removes all unknown bases from the file
  group_by(Group,SNP_ID) %>%  #Grouping by Group and SNP_ID
  filter(length(unique(Base))>1) %>%   #This removes all SNPs with one base
  select(Group, SNP_ID, Chromosome) #Selecting three fields

Group_graph<-Group_graph[!duplicated(Group_graph),] #Getting rid of duplication

ggplot(data=Group_graph) +
  geom_bar(mapping=aes(x=Chromosome, fill=Group)) + 
  scale_x_discrete(limit=c(1:10), label=c(1:10))

```
Converting the data into long format

```{r}
#New column is created on the basis of homozygosity.

Pivot$homozygous<-"Heterozygous"
Pivot$homozygous[Pivot$Base=="?/?"]<-"Missing"
Pivot$homozygous[Pivot$Base %in% c("A/A","C/C", "G/G", "T/T")]<-"Homozygous"

# SNPs are grpahed by Sample_ID, filled by Homozygosity:
ggplot(data=Pivot) +
  geom_bar(mapping=aes(x=Sample_ID, fill=homozygous), position="fill") 
# Graph the SNPs by Group, filling by Homozygosity:
ggplot(data=Pivot) +
  geom_bar(mapping=aes(x=Group, fill=homozygous), position="fill")

```

To see polymorphism in each group in each chromosome. 
```{r, fig.width=10,fig.height=11}

Poly<-Pivot[!is.na(as.numeric(Pivot$Chromosome)),]
Poly$Chromosome<-as.numeric(Poly$Chromosome)
Poly<-Poly%>% select(SNP_ID, Group, Base, Position, Chromosome, homozygous)%>% filter(Base!="?/?")%>% unique()%>%filter(!is.na(as.numeric(Position)))
Graph<-ggplot(data = Poly) + geom_point(mapping=aes(x=as.numeric(Position), y=Group, color=homozygous)) +labs(y = "Groups" , x="Chromosome position")

Graph + facet_wrap(~ Chromosome,ncol=2,scales = "free") + labs(color="Genotype")
```
